---
title: "gggibbous: Moon charts, a pie chart alternative for two groups"
author: "Mike Bramson"
date: "October 27, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gggibbous}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
img {
  border: none;
  display: block;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

required <- c("maps", "mapproj")
if (!all(sapply(required, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

## Moon charts and pie charts

A pie chart divides a circle into multiple sections where the arc lengths (and
so also the areas) of the sections represent proportions of a whole.
A moon chart, similarly, divides a circle into sections where the areas
represent proportions of a whole, but in a moon chart the areas are drawn as
crescent or gibbous portions of a circle—like the phases of the moon.

```{r, echo = FALSE, fig.width = 3, fig.height = 1.5}
ggplot2::ggplot(
  data.frame(
    y = c(1, 3), group = c("a", "b"),
    facet = factor(c(rep("pie", 2)), levels = c("pie", "moon"))
  ),
  ggplot2::aes(y = y, fill = group)
) +
  ggplot2::geom_col(ggplot2::aes(x = factor(0)), width = 2, color = "black") +
  gggibbous::geom_moon(
    data = data.frame(
      x = factor(rep(-0.5, 2)), y = rep(0, 2), ratio = c(0.75, 0.25),
      right = c(TRUE, FALSE), group = c("b", "a"), facet = c("moon", "moon")
    ),
    ggplot2::aes(x = x, y = y, ratio = ratio, right = right),
    size = 36, stroke = 0.5
  ) +
  ggplot2::coord_polar(theta = "y") +
  ggplot2::facet_wrap(~facet) +
  ggplot2::scale_fill_manual(values = c("white", "black"), guide = "none") +
  ggplot2::theme_void(16)
```

The motivation behind using a moon chart instead of a pie chart is primarily
one of aesthetic choice.
Note also that because the sections of a moon chart are swept from one or the
other side of the circle, they are generally only appropriate for depicting one
or two groups.

## `gggibbous` and its usage

`gggibbous` extends the `ggplot2` data visualization package to provide support
for moon charts in R.
Unlike the pie charts supported natively by `coord_polar()` in R, moon charts in
`gggibbous` do not require any special coordinate system.
They are drawn most similarly to points in `ggplot2`: their position is defined
by an `x` and a `y` coordinate and their size is defined independently of the
coordinate system, so they always remain circular.

```{r setup}
library(gggibbous)
```

```{r, fig.width = 4, fig.height = 2}
ggplot(data.frame(x = 1:5, y = 1, size = 2^(0:4)), aes(x, y, size = size)) +
  geom_moon() +
  geom_point(y = 2) +
  lims(x = c(0.5, 5.5), y = c(0.5, 2.5)) +
  scale_size(range = c(5, 10))
```

Two new aesthetics are also important in `geom_moon`: `ratio` and `right`.

### The `ratio` aesthetic

`ratio` controls the proportion of the moon to be drawn.
It should be between 0 (a “new moon” where nothing is actually drawn) and 1 (a
“full moon”, i.e. a circle).

```{r, fig.width = 5, fig.height = 1.25}
ggplot(data.frame(x = 1:5, y = 0, ratio = 0:4 * 0.25), aes(x = x, y = y)) +
  geom_moon(aes(ratio = ratio), size = 20, fill = "black") +
  geom_text(aes(y = y + 1, label = ratio)) +
  lims(x = c(0.5, 5.5), y = c(-1, 1.4)) +
  theme_void()
```

### The `right` aesthetic

`right` takes a boolean value that controls whether the moon is “waxing” or
“waning”—that is, whether it is “filled” from the right or the left.

One way to make a “complete” moon with two colors is to use `right = TRUE` for
one color and `right = FALSE` for the other, with complementary ratios.
(See the examples below for some other approaches.)

```{r, fig.width = 3.5, fig.height = 2.5}
tidymoons <- data.frame(
  x = rep(1:3, 6),
  y = rep(rep(3:1, each = 3), 2),
  ratio = c(1:9 / 10, 9:1 / 10),
  right = rep(c(TRUE, FALSE), each = 9)
)

ggplot(tidymoons) +
  geom_moon(aes(x, y, ratio = ratio, right = right, fill = right)) +
  lims(x = c(0.5, 3.5), y = c(0.5, 3.5))
```

## Legend key glyphs

`gggibbous` includes three key glyphs for different types of legends:
`draw_key_moon`, `draw_key_moon_left`, and `draw_key_full_moon`

**`draw_key_moon`**—the default in `geom_moon`—draws a gibbous moon (see above).

**`draw_key_moon_left`** draws a crescent moon from the left that is
complementary to the gibbous moon in `draw_key_moon`, which is useful for
combined legends:

```{r, fig.width = 3.5, fig.height = 2.5}
ggplot(tidymoons, aes(x, y, ratio = ratio, right = right, size = 2^x)) +
  geom_moon(data = subset(tidymoons, right), fill = "violetred") +
  geom_moon(
    data = subset(tidymoons, !right), fill = "turquoise3",
    key_glyph = draw_key_moon_left
  ) +
  lims(x = c(0.5, 3.5), y = c(0.5, 3.5)) +
  scale_size("size", range = c(5, 10), breaks = 2^(1:3))
```

**`draw_key_full_moon`** draws a circle.
It is similar to the “point” key glyph, but the size is calculated slightly
differently, so it is more appropriate if you want the size of legend moons
and the size of the plot moons to match.

```{r, fig.width = 4.5, fig.height = 2.5}
ggplot(tidymoons) +
  geom_moon(
    aes(x, y, ratio = ratio, right = right, fill = right, size = 2^x),
    key_glyph = draw_key_full_moon
  ) +
  lims(x = c(0.5, 3.5), y = c(0.5, 3.5)) +
  scale_size("size", range = c(5, 10), breaks = 2^(1:3)) +
  scale_fill_manual(values = c("firebrick1", "dodgerblue2")) +
  theme(legend.box = "horizontal")
```

## Worked examples

### Moon charts on maps

One common use of multiple pie charts is to represent proportions at different
coordinates on a map.
The *x* and *y* dimensions are already committed to the map coordinates, so
proportional visualizations like bar charts are more difficult.
This is a perfect opportunity to try out moon charts!

Pie chart maps are popular in population genetics, so let's look at an example
from that field.
The `dmeladh` data^[Oakeshott, J.G., et al. 1982. Alcohol dehydrogenase and
glycerol-3-phosphate dehydrogenase clines in *Drosophila melanogaster* on
different continents. Evolution, 36(1): 86-96.] contains frequencies of two
variants of the *Adh* gene in Australian and Papua New Guinean fruit fly
populations.
Many of these populations are close together, so we have to deal with
overplotting, which we do manually below.

```{r, fig.width = 6, fig.height = 5.45}
dmeladh_adj <- dmeladh
dmeladh_adj$long <- dmeladh$Longitude + c(
  -2, 0, -2, 2, -3, 3, 3, 2, 3, 4, -2.5, -2.5, -1, -2, -2.5, -4, 2.5,
  5, 6, 7, 2, -7, -5.5, -3, 0, -7, -2, 3, 5.5, 0.5, -1, -1.5, -3, 2)
dmeladh_adj$lat <- dmeladh$Latitude + c(
  -2, 2, 0, 1, 0, 0, 0, 2, 0.5, -1, 1, -1.5, 2, 4, 1.5, 0, 2,
  1, -1, -3, -2, 1, -1, -2, -3, -2, -4, -3, -1, 1.5, 2, 2, -2, 0)

moonmap <- ggplot(dmeladh_adj, aes(long, lat)) +
  geom_polygon(
    data = map_data(
      "world", region = "(Australia)|(Indonesia)|(Papua New Guinea)"),
    aes(group = group),
    fill = "gray80"
  ) +
  geom_segment(aes(xend = Longitude, yend = Latitude), color = "gray20") +
  geom_point(aes(Longitude, Latitude), size = 0.75, color = "gray20") +
  scale_size(range = c(4, 10)) +
  coord_map(xlim = c(110, 160), ylim = c(-45, -5)) +
  theme_void() +
  theme(
    legend.position = c(0.05, 0.05),
    legend.direction = "horizontal",
    legend.justification = c(0, 0)
  )

moonmap +
  geom_moon(
    aes(ratio = AdhS / 100, size = N),
    right = FALSE, fill = "gold", color = "gold",
    key_glyph = draw_key_moon_left
  ) +
  geom_moon(
    aes(ratio = AdhF / 100, size = N),
    fill = "forestgreen", color = "forestgreen"
  )
```

If we want to label the alleles in the legend explicitly, then we need to map
them to a group, which requires that we rearrange the data into a “longer”
(“tidy”) format.

```{r, fig.width = 6, fig.height = 5.45}
adhf <- dmeladh_adj[c("long", "lat", "N", "AdhF")]
adhs <- dmeladh_adj[c("long", "lat", "N", "AdhS")]
names(adhf)[4] <- "percent"
names(adhs)[4] <- "percent"
tidyadh <- rbind(adhf, adhs)
tidyadh$allele <- rep(c("AdhF", "AdhS"), each = nrow(dmeladh_adj))
tidyadh$right <- rep(c(TRUE, FALSE), each = nrow(dmeladh_adj))

moonmap +
  geom_moon(
    data = tidyadh, key_glyph = draw_key_full_moon,
    aes(ratio = percent / 100, fill = allele, color = allele, right = right,
        size = N)
  ) +
  scale_fill_manual(values = c("forestgreen", "gold")) +
  scale_color_manual(values = c("forestgreen", "gold"))
```

### Lunar data

Sometimes you just want to plot literal representations of the moon.
The `lunardist` data, adapted from NASA, contains the distance from the Earth to
the Moon for each day in 2019, as well as the dates (in UTC) of each occurrence
of the four principal phases of the moon.
We can plot those principal phases using moon charts (which in this case are
identical to pie charts).

```{r, fig.width = 7, fig.height = 3.5}
moonphase <- subset(lunardist, !is.na(phase))
moonphase$percent <- ifelse(
  moonphase$phase == "new", 0, ifelse(moonphase$phase == "full", 1, 0.5))

ggplot(lunardist, aes(date, distance)) +
  geom_line() +
  # Plotting the lower layer as a full circle also works in most cases
  geom_moon(data = moonphase, ratio = 1, size = 5, fill = "black") +
  geom_moon(
    data = moonphase, aes(ratio = percent),
    size = 5, fill = "yellow", right = moonphase$phase == "first quarter"
  )
```

TODO:

* Harvey Ball example